Windowed least-squares model reduction for dynamical systems

### 2020-10-12 numerical_experiments.tex
latest

\section{Numerical experiments}\label{sec:numerical_experiments}
We now analyze the performance of \methodAcronymROMs\ leveraging \spatialAcronym\ trial subspaces on two benchmark problems: the Sod shock tube and compressible flow in a cavity. In each experiment, we compare \methodAcronymROMs\ to the Galerkin and LSPG ROMs. The purpose of the numerical experiments is to assess the impact of minimizing the residual over an arbitrarily sized time window on the solution accuracy. We additionally assess the impact of the time step and time scheme on \methodAcronym. In both experiments, the spatial basis is equivalent for each time window, e.g., $\basisspaceArg{n} \equiv \basisspace$, $n=1,\ldots,\nslabs$. We also note that both experiments are designed to test the \textit{reproductive} ability of the ROMs. We do not consider future state prediction and prediction at new parameter instances as these problems introduce factors that confound the solution accuracy with the solution methodology (e.g., accuracy of the basis).

@@

\subsubsection{Description of FOM and generation of \spatialAcronym\ trial subspace}\label{sec:sod_fom}
We solve the 1D compressible Euler equations with a finite volume method. We partition the domain into 500 cells of uniform width and employ the Rusanov flux~\cite{rusanov} at the cell interfaces. We employ the Crank--Nicolson (CN) scheme, which is a linear multistep method defined by the coefficients $\alpha_0 = 1,\alpha_1 = -1, \beta_0 = \beta_1 = 1/2$, for temporal integration. We evolve the FOM for $t \in [0.0,1.0]$ at a time-step of $\Delta t = 0.002$. We construct the \spatialAcronym\ trial subspace by executing Algorithm~\ref{alg:pod} with inputs $N_{\text{skip}}=2, \stateIntercept = \bz, K=46 $, and $\mathbf{P} = \mathbf{I}$. The resulting trial subspace corresponds to an energy criterion of $99.99\%$.
%  The solution is saved every other time-step, resulting in 500 solutions snapshots. These snapshots are used to generate a trial basis for the reduced-order models through POD. The trial basis is selected to be of dimension $K = 46$, which corresponds to a $99.99\%$ energy criterion. No affine transformation is used in the construction of the trial subspace, i.e., $\stateIntercept = \bz$.

\subsubsection{Description of reduced-order models}
We consider reduced-order models based on Galerkin projection, LSPG
projection, and the \methodAcronym\ approach. No hyper-reduction is considered 
in this example, i.e., $\stweightingMatArg{n} = \mathbf{I}$, $n=1,\ldots,\nslabs$. Details on the implementation of the 
different reduced-order models is as follows:
\begin{itemize}
\item \textit{Galerkin ROM}: We obtain the Galerkin ROM through Galerkin projection of the FOM and evolve the Galerkin ROM in time with the CN time scheme at a constant time step of $ \Delta t = 0.002$.

@@

Next, Figure~\ref{fig:sod_error} shows time-integrated $\ell^2$-norm of the state errors and the objective function~\eqref{eq:obj} (i.e., the time-integrated residual norm) over $t \in [0,1]$ for the various ROMs. Most notably, we observe 
that increasing the window size over which the residual is minimized does \textit{not} lead to a monotonic decrease in the error as measured in the time-integrated $\elltwo$-norm (we refer to this as the $\elltwo$-error). As expected, increasing the window size does lead to a monotonic decrease in the time-integrated residual norm, however. We additionally note that, although the time-integrated $\elltwo$-error of 
the projected FOM solution is significantly lower than that of the various ROM solutions, the time-integrated residual norm of the projected FOM solution is \textit{higher} 
than all ROMs. Thus, although the projected FOM solution is more accurate in the $\elltwo$-norm, it does not satisfy the governing equations as well the ROM solutions.

@@

\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_converge.pdf}
\caption{Relative time-integrated $\elltwo$-error of the \methodAcronymROM.}
\label{fig:sod_error_converge}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/residual_converge.pdf}
\caption{Time-integrated residual norm of the \methodAcronymROM.} 
\label{fig:sod_residual_converge}
\end{subfigure}
\caption{Performance metrics of the Galerkin, LSPG, and \methodAcronymROMs\ as a function of basis dimension. Missing entries indicate that the solver did not converge.} 
\label{fig:convergence_study}
\end{center}
\end{figure}

@@

\input{cavity}

>>>>

\section{Numerical experiments}\label{sec:numerical_experiments}
We now analyze the performance of \methodAcronymROMs\ leveraging \spatialAcronym\ trial subspaces on three benchmark problems: (1) a reproductive ROM for the Sod shock tube problem (2) a reproductive ROM for a supersonic compressible flow in a cavity, and (3) a predictive ROM for the parameterized shallow water equations. In each experiment, we compare \methodAcronymROMs\ to either the LSPG, or the Galerkin and LSPG, ROMs. In the experiments, we assess (1) the impact of minimizing the residual over an arbitrarily sized time window on the solution accuracy, (2) the impact of the time step and time scheme on \methodAcronym, (3) the impact of the window size on the convergence properties of the ROMs, and (4) the impact of the window size on the cost of the ROMs. In all experiments, the spatial basis is equivalent for each time window, e.g., $\basisspaceArg{n} \equiv \basisspace$, $n=1,\ldots,\nslabs$. We note that the first two experiments are designed to test the \textit{reproductive} ability of the ROMs, while the third experiment tests the predictive capabilities of the ROM.

@@

\subsubsection{Description of FOM and generation of \spatialAcronym\ trial subspace}\label{sec:sod_fom}
We solve the 1D compressible Euler equations with a finite volume method. We partition the domain into 500 cells of uniform width and employ the Rusanov flux~\cite{rusanov} at the cell interfaces. We employ the Crank--Nicolson (CN) scheme, which is a linear multistep method defined by the coefficients $\alpha_0 = 1,\alpha_1 = -1, \beta_0 = \beta_1 = 1/2$, for temporal integration. We evolve the FOM for $t \in [0.0,1.0]$ at a time-step of $\Delta t = 0.002$. We construct the \spatialAcronym\ trial subspace by executing Algorithm~\ref{alg:pod} with inputs $N_{\text{skip}}=2, \stateIntercept = \bz, K=46 $, and $\mathbf{T} = \mathbf{I}$. The resulting trial subspace corresponds to an energy criterion of $99.99\%$.
%  The solution is saved every other time-step, resulting in 500 solutions snapshots. These snapshots are used to generate a trial basis for the reduced-order models through POD. The trial basis is selected to be of dimension $K = 46$, which corresponds to a $99.99\%$ energy criterion. No affine transformation is used in the construction of the trial subspace, i.e., $\stateIntercept = \bz$.

\subsubsection{Description of reduced-order models}\label{sec:sod_rom_description}
We consider reduced-order models based on Galerkin projection, LSPG
projection, and the \methodAcronym\ approach. No hyper-reduction is considered 
in this example, i.e., $\stweightingMatArg{n} = \mathbf{I}$, $n=1,\ldots,\nslabs$. Details on the implementation of the 
different reduced-order models is as follows:
\begin{itemize}
\item \textit{Galerkin ROM}: We obtain the Galerkin ROM through Galerkin projection of the FOM and evolve the Galerkin ROM in time with the CN time scheme at a constant time step of $ \Delta t = 0.002$.

@@

Next, Figure~\ref{fig:sod_error} shows time-integrated $\ell^2$-norm of the state errors and the objective function~\eqref{eq:obj} (i.e., the time-integrated residual norm) over $t \in [0,1]$ for the various ROMs. Most notably, we observe 
that increasing the window size over which the residual is minimized does \textit{not} lead to a monotonic decrease in the error as measured in the time-integrated $\elltwo$-norm (we refer to this as the $\elltwo$-error). As expected, increasing the window size does lead to a monotonic decrease in the time-integrated residual norm, however. We additionally note that, although the time-integrated $\elltwo$-error of 
the projected FOM solution is significantly lower than that of the various ROM solutions, the time-integrated residual norm of the projected FOM solution (i.e., the time-integrated norm of the residual evaluated at the $\elltwo$-orthogonal projection of the FOM solution onto the trial subspace) is \textit{higher} 
than all ROMs. Thus, although the projected FOM solution is more accurate in the $\elltwo$-norm, it does not satisfy the governing equations as well the ROM solutions.

@@

\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_converge.pdf}
\caption{Relative time-integrated $\elltwo$-error.}
\label{fig:sod_error_converge}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/residual_converge.pdf}
\caption{Time-integrated residual norm.} 
\label{fig:sod_residual_converge}
\end{subfigure}
\caption{Performance metrics of the Galerkin, LSPG, and \methodAcronymROMs\ as a function of basis dimension. Missing entries indicate that the solver did not converge.} 
\label{fig:convergence_study}
\end{center}
\end{figure}

@@

\input{cavity}
\input{shallow_water}

### 2020-10-12 section4/spatial_numerical.tex
latest

\begin{remark}\label{remark:gaussnewton}\textit{(Acceleration of the
	Gauss--Newton Method)}\\
	The principal cost of a Gauss--Newton method is often the formation of the Jacobian matrix. A variety of techniques aimed at 
reducing this computational burden exist; Jacobian-free Newton--Krylov
	methods~\cite{jfnk}, Broyden's method~\cite{broyden} (as explored in
	Ref.~\cite{carlberg_thesis}, Appendix A), and frozen Jacobian approximations
	are several such examples. Further, the space--time formulation introduces
	an extra dimension for parallelization that can be exploited to future
	reduce the wall time. The investigation of these additional, potentially
	more efficient, solution algorithms is a topic of future work. 
\end{remark}
\begin{algorithm}
\caption{\spatialAcronym\ trial subspace: algorithm for the direct solution technique with the Gauss--Newton method and a linear multistep method over the $n$th window}
\label{alg:colloc_gn}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\epsilon$; initial guess, $\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0}$}%; initial condition $\genstateDiscreteArg{n,0}$}
\Output{Solution to least squares problem, $\genstatecollocMatSlabArg{n}$} 
\textbf{Online Steps}: \\
$\text{converged} \leftarrow \text{false}$ \Comment{Set convergence checker} \\
$\genstatecollocMatSlabArg{n}_0 \leftarrow \unroll(\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0})$ \Comment{Assemble generalized coordinates over window} \\
$k \leftarrow 0$ \Comment{Set counter}\\
\While{\text{not converged}}
{
%\For{$i=1,\hdots,\nstepsArg{n}$}{
%  \For{$j=1,\hdots,\ncollocArg{n}{i}$}{
%Compute: $\approxstateArgnt{n,i}  =  \basisspace  \genstateDiscreteArg{n,i} + \stateIntercept$ 
%\Comment{Compute state} \\
%Compute: $\velocity(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept ) $ \Comment{Compute velocity at each time-step instance}\\
%Compute: $\residLMSArg{n,i}(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept, \ldots , \basisspace \genstateDiscreteArg{n,i - k^n(i)} + \stateIntercept) $  \Comment{Compute residual} \\
%}
$\mathbf{r} \leftarrow \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{Compute weighted residual over window} \\
$\mathbf{J} \leftarrow  
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n}_k) 
$ \Comment{Compute weighted residual-Jacobian over window} \\
%Compute: $[\jacobianSlabArg{n}]^T \jacobianSlabArg{n}$ \Comment{Compute system matrix for the normal equations} \\
%Compute: $[\jacobianSlabArg{n}]^T \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{compute RHS for normal equations} \\
\uIf{ $\norm{ \mathbf{J}^T\mathbf{r}  } \le \epsilon$ }{
{\text{converged} $\leftarrow$ \text{true}}  \Comment{Check and set convergence based on gradient norm} \\
Return: $\genstatecollocMatSlabArg{n} = \genstatecollocMatSlabArg{n}_{k+1} $ \Comment{Return converged solution}\\
}
\Else{
 Compute  $\Delta  \genstatecollocMatSlabArg{n} $ that minimizes $\norm{ \mathbf{J} \Delta \genstatecollocMatSlabArg{n} +\mathbf{r}}^2$ \Comment{Solve the linear least-squares problem} \\
$\alpha \leftarrow \text{linesearch}(\Delta  \genstatecollocMatSlabArg{n} ,\genstatecollocMatSlabArg{n}_k ) $ \Comment{Compute $\alpha$ based on a line search, or set to 1}\\
$\genstatecollocMatSlabArg{n}_{k+1} \leftarrow \genstatecollocMatSlabArg{n}_k + \alpha \Delta \genstatecollocMatSlabArg{n}$ \Comment{Update guess to the state} \\
}
$k\leftarrow k+1$
}
\end{algorithm}

>>>>

\begin{remark}\label{remark:gaussnewton}\textit{(Acceleration of the
	Gauss--Newton Method)}\\
	The principal cost of a Gauss--Newton method is often the formation of the Jacobian matrix. A variety of techniques aimed at 
reducing this computational burden exist; Jacobian-free Newton--Krylov
	methods~\cite{jfnk}, Broyden's method~\cite{broyden} (as explored in
	Ref.~\cite{carlberg_thesis}, Appendix A), and frozen Jacobian approximations
	are several such examples. Further, the space--time formulation introduces
	an extra dimension for parallelization that can be exploited to future
	reduce the wall time. While we leave a detailed investigation of these additional, potentially
	more efficient, solution algorithms as a topic of future work, we briefly examine the use of frozen Jacobians in Section~\ref{sec:swe} and Appendix~\ref{appendix:gnupdate}. 
\end{remark}
\begin{algorithm}
\caption{\spatialAcronym\ trial subspace: algorithm for the direct solution technique with the Gauss--Newton method and a linear multistep method over the $n$th window}
\label{alg:colloc_gn}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\epsilon$; initial guess, $\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0}$}%; initial condition $\genstateDiscreteArg{n,0}$}
\Output{Solution to least squares problem, $\genstatecollocMatSlabArg{n}$} 
\textbf{Online Steps}: \\
$\text{converged} \leftarrow \text{false}$ \Comment{Set convergence checker} \\
$\genstatecollocMatSlabArg{n}_0 \leftarrow \unroll(\genstateGuessDiscreteArg{n,1}{0},\ldots,\genstateGuessDiscreteArg{n,\nstepsArg{n}}{0})$ \Comment{Assemble generalized coordinates over window} \\
$k \leftarrow 0$ \Comment{Set counter}\\
\While{\text{not converged}}
{
%\For{$i=1,\hdots,\nstepsArg{n}$}{
%  \For{$j=1,\hdots,\ncollocArg{n}{i}$}{
%Compute: $\approxstateArgnt{n,i}  =  \basisspace  \genstateDiscreteArg{n,i} + \stateIntercept$ 
%\Comment{Compute state} \\
%Compute: $\velocity(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept ) $ \Comment{Compute velocity at each time-step instance}\\
%Compute: $\residLMSArg{n,i}(\basisspace \genstateDiscreteArg{n,i} + \stateIntercept, \ldots , \basisspace \genstateDiscreteArg{n,i - k^n(i)} + \stateIntercept) $  \Comment{Compute residual} \\
%}
$\mathbf{r} \leftarrow \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{Compute weighted residual over window} \\
$\mathbf{J} \leftarrow  
\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}(\genstatecollocMatSlabArg{n}_k) 
$ \Comment{Compute weighted residual-Jacobian over window} \\
%Compute: $[\jacobianSlabArg{n}]^T \jacobianSlabArg{n}$ \Comment{Compute system matrix for the normal equations} \\
%Compute: $[\jacobianSlabArg{n}]^T \residLMSSlabArg{n}(\genstatecollocMatSlabArg{n}_k)$ \Comment{compute RHS for normal equations} \\
\uIf{ $\norm{ \mathbf{J}^T\mathbf{r}  } \le \epsilon$ }{
{\text{converged} $\leftarrow$ \text{true}}  \Comment{Check and set convergence based on gradient norm} \\
Return: $\genstatecollocMatSlabArg{n} = \genstatecollocMatSlabArg{n}_{k+1} $ \Comment{Return converged solution}\\
}
\Else{
 Compute  $\Delta  \genstatecollocMatSlabArg{n} $ that minimizes $\norm{ \mathbf{J} \Delta \genstatecollocMatSlabArg{n} +\mathbf{r}}^2$ \Comment{Solve the linear least-squares problem} \\
$\alpha \leftarrow \text{linesearch}(\Delta  \genstatecollocMatSlabArg{n} ,\genstatecollocMatSlabArg{n}_k ) $ \Comment{Compute $\alpha$ based on a line search, or set to 1}\\
$\genstatecollocMatSlabArg{n}_{k+1} \leftarrow \genstatecollocMatSlabArg{n}_k + \alpha \Delta \genstatecollocMatSlabArg{n}$ \Comment{Update guess to the state} \\
}
$k\leftarrow k+1$
}
\end{algorithm}
